Choosing a Population Genomics Approach

### 2021-11-04 RADseq.Rmd
Changed format of site to Bootswatch

---
layout: default
title: RADseq
parent: Choosing a PopGenomics Approach
grand_parent: Best Principles
#editor_options: 
#  chunk_output_type: console
---

@@

#set bash code chunks to use bash_profile
knitr::opts_chunk$set(engine.opts = list(bash = "-l"))

if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

#BiocManager::install("SeqArray")
#BiocManager::install("SNPRelate")
#install.packages("rlang")
```

# RADseq  
{: .no_toc }

<details open markdown="block">
  <summary>
    Table of contents
  </summary>
  {: .text-delta }
1. TOC
{:toc}
</details>

If you want to run the acompanying code chunks, install Docker container.
```{r}
library(ggplot2)
library(SeqArray)
library(tidyverse)
```

<!-- `r colorize("Dani to do: make a dockerfile with all the packages used in this Rscript for reproducibility", "red")` -->

<!-- `r colorize("Example files needed: example strata file 'data/example.metadata.csv', example vcf for PCA './data/test.vcf' ", "red")` -->

# Reduced Representation Sequencing (RADseq/GBS)  
<!-- `r colorize("What kinds of protocols is this advice relevant to?", "red")`   -->
"Restriction-site   Associated   DNA   sequencing"  - RADseq - combines  restriction  enzymes  with next-gen, massively parallel,  short-read  sequencing. RADseq involves the use of restriction enzymes which are used to shear DNA at selected sites. RADseq comes in different flavors. Double-digest RADseq (ddRAD; Peterson et al.,2012)  selects  markers  with  two  restriction enzymes  with  different  cut  frequencies. Genotype-by-Sequencing  (GBS;  Elshire  et  al.,  2011)  uses  a  frequent-cutting  restriction  enzyme  with PCR size selection. See this review for a comparison of some of the more popular protocols (link). In this guide we use the term "RADseq" to refer to any of these protocols, including those that don't involve random shearing of data. When recommendations are specific to a certain type of RAD/GBS, we will explicitly say so.

<!-- `r colorize("characteristics of radseq that are different/ important to know compared to other types of sequencing ", "red")`   -->

@@

![Table 1 from O'Leary et al. 2018.](../RAD_code/images/oleary_table1.png)

@@

* If this RAD/GBS method has not been done in your species or in your molecular lab setup before, spend some time optimizing the protocol using a representative subset of individuals. Then try to keep everything about the library prep as consistent as possible across samples (eg, DNA extraction kit, PCR cycles, sequencing platform). This isn't always possible, especially if you need to optimize the protocol for certain tricky samples. Just make sure to keep track of everything!   
* Randomize samples across library prep batches and sequencing lanes! For example, if you are sequencing two different groups of samples on two different sequencing lanes, make sure they are randomized in respect to sample location or whatever your groups of interest are. If you are preparing groups of individuals in different batches to be pooled later, randomize across these batches. Alicia MY even recommends randomizing across DNA extractions, especially if the person doing the extractions is new to molecular work. This is to allow you to control for potential [batch effects](#batch) that are often observed with RAD data.  
```{r}
# This code reads in a comma-delimited STRATA/meta-data file, randomises samples, and returns a file with wells and plates and duplicates which may be used to assist with plating libraries
# ** TODO** will need to consider where to duplicate samples - need to consider STARTA, DOC, ect..?
# maybe have user write in a list of samples to replicate

@@

### First, look at the raw data! {#fastqc}  
Always look at your data with fastqc(link) before starting an assembly. First, this is a good check to just make sure the sequencing worked. If you have demultiplexed data, you can use multiqc(link) to generate fastqc plots for all individuals and quickly identify ones that did not sequence well.  
Check out [this informative tutorial](https://datacarpentry.org/wrangling-genomics/02-quality-control/index.html) on running fastqc and interpreting the output.  
Do you have a lot of adapter sequences? Are the ends of your reads very low quality? If so, you should expect a fair amount of trimming and read filtering to occur prior to assembly. If that doesn't occur or if too many reads are being filtered so as to only recover a small number of SNPs, something might need to be tweaked with your trimming and filtering step. [See how to evaluate for sequencing error](#seq)

@@

@@

@@

One of the most powerful methods for exploring your data for issues is a Principle Components Analysis (see [here](https://michaelaalcorn.medium.com/yet-another-pca-explanation-727dff6ce26a) for a linear algebr-based explanation of PCA). From [this genetics tutorial](https://comppopgenworkshop2019.readthedocs.io/en/latest/contents/03_pca/pca.html): "To understand how PCA works, consider a single individual and its representation by its 593,124 markers. Formally, each individual is a point in a 593,124-dimensional space, where each dimension can take only the three possible genotypes indicated above, or have missing data. To visualise this high-dimensional dataset, we would like to project it down to two dimensions. But as there are many ways to project the shadow of a three-dimensional object on a two dimensional plane, there are many (and even more) ways to project a 593,124-dimensional cloud of points to two dimensions. What PCA does is figuring out the “best” way to do this project in order to visualise the major components of variance in the data.

As soon as you have a file with SNPs for all samples, before extensive filtering or playing with parameters, do a PCA on it and look at all the axes with the most variation. Be careful that you understand how missing data is imputed for PCA.

For the code examples in this section, we mostly make use of the R package SeqArray, which can read in and manipulate VCFs (*notes on Docker install*). We like this program because .... For those you are less familiar with R or more familar with Plink, check out [this tutorial]() which uses Plink and minimal R plotting.  a tutorial that uses We also include some examples using VCFtools and minimal R plotting, for those who are less familiar with R.

(?) dani asks: you want me to make this take multiple different vcfs from pipeline outputs for plotting? should we implement pcANGSD which can handle missing data better  (?)  
K: Making it take multiple vcfs is a secondary goal, and can be added later if we have time/interest. Adding a code example for pcANGSD would be useful though.

## ill get onto this sorry

Doing a PCA with SeqArray and SNPRelate in R. Filtering for SNPs with a MSF > 2.5%
```{r, eval =T}
## read in VCF and convert to GDS format
#vcf.file = "./data/test.vcf"
#seqVCF2GDS(vcf.file, "test.gds.rad", storage.option="ZIP_RA")

# read in existing gds file  
filename = "data/test.gds.rad"
gdsin= SeqArray::seqOpen(filename,)
# summarize the data
print(paste0("The number of SAMPLES in data: ", length(c(SeqArray::seqGetData(gdsin, "sample.id")))))
print(paste0("The number of SNPs in data: ",  length(c(SeqArray::seqGetData(gdsin, "variant.id")))))
#read in metadata
sample.ids = seqGetData(gdsin, "sample.id")
sample.strata =  read.table("data/test.metadata.csv", header = T, sep = "\t") %>%
                  dplyr::select(TARGET_ID, STRATA, LONG, LAT, PLATE)
id.order = sapply(sample.ids, function(x,df){which(df$TARGET_ID == x)}, df=sample.strata) #incase your strata file is not in the same order as your vcf
sample.strata.order = sample.strata[id.order,]

@@

@@

```{r}
#look at PC values for reps
#remove reps
```

First color/shape the individuals by the factor you expect to matter (eg, sampling site or region, family, ecotype).  
```{r, eval=T}
#PLOT PCA
#PC1 v PC2 colored by collection location
print(
  as.data.frame(pca.out$eigenvect) %>%
      #tibble::add_column(., ID =  sample.ids) %>%
      tibble::add_column(., STRATA =  sample.strata.order$STRATA) %>%
      ggplot(., aes(x=V1, y=V2, color = STRATA))  + #label = ID
      geom_point(size=2) +
      #geom_text() +
      stat_ellipse(level = 0.95, size = 1) +
      #scale_color_manual(values = gds_colors) +
      geom_hline(yintercept = 0) +
      geom_vline(xintercept = 0) +
      theme_bw() +
      xlab(paste0("PC1 [",paste0(round(eig[1], 2)), "%]")) +
      ylab(paste0("PC2 [",paste0(round(eig[2], 2)), "%]")) +
      ggtitle("PCA Colored by Collection Location")
)
```

* Is there any clustering? If so, is it different than your expectation?
  + If different than expectation, suggests either [batch effects](#batch), [cryptic variation](#cryptic) among individuals, or systematic biological issues affecting assembly/mapping (eg, inversions (Katies inversion, also paper ref))  
  <!-- + `r colorize("figure example, weird clustering", "red")` -->
* Are there outlier samples driving a large amount of variation?
  + If so, they may be [cryptic species](#cryptic), clones/sample replicates, or they [sequenced/genotyped poorly](#bad)
  + In our example we see 2 potential outlier samples in the bottom left corner. We can check which samples they are by plotting the PCA with sample ID and/or looking at the actual PC scores.
```{r, eval=T}
#PLOT PCA
#PC1 v PC2 colored by collection location
print(
  as.data.frame(pca.out$eigenvect) %>%
      tibble::add_column(., ID =  sample.ids) %>%
      ggplot(., aes(x=V1, y=V2, label = ID)) +
      geom_text(size =1.5) +
      geom_hline(yintercept = 0) +
      geom_vline(xintercept = 0) +
      theme_bw() +
      xlab(paste0("PC1 [",paste0(round(eig[1], 2)), "%]")) +
      ylab(paste0("PC2 [",paste0(round(eig[2], 2)), "%]")) +
      ggtitle("PCA with Labels")
)
```
```{r}
# which samples have PC1 < -0.1 and PC2 < -0.6?
as.data.frame(pca.out$eigenvect) %>%
      tibble::add_column(., ID =  sample.ids) %>%
  filter(V1 < -0.1 & V2 < -0.6) %>%
  select(ID,V1,V2)
```

* Is there no clustering at all? If so, it may be due to:  
  + bioinformatic artifacts leading to noise, [see Testing Parameters](#test)
  + issues with how you are [filtering the data](#filter) (eg, too strict or too lax)
  + issues with how your PCA is treating missing data
  + or you actually have no structure in your dataset!
  <!-- + `r colorize("figure example, no clustering", "red")` -->

Next, color the individuals by potential sources of batch effects (sequencing lane, library prep batch, age of tissue sample, person doing the DNA extraction...).    
```{r, eval=T}
# again, but colored by batch
print(
  as.data.frame(pca.out$eigenvect) %>%
      tibble::add_column(., PLATE =  as.factor(sample.strata.order$PLATE)) %>%
      ggplot(., aes(x=V1, y=V2, color = PLATE))  + #label = ID
      geom_point(size=2) +
      stat_ellipse(level = 0.95, size = 1) +
      geom_hline(yintercept = 0) +
      geom_vline(xintercept = 0) +
      theme_bw() +
      xlab(paste0("PC1 [",paste0(round(eig[1], 2)), "%]")) +
      ylab(paste0("PC2 [",paste0(round(eig[2], 2)), "%]")) +
      ggtitle("PCA colored by Batch/Sequencing Plate")
)
```

@@

#### Visually inspect alignments {#align}  
Another useful approach is to visually inspect alignments for SNPs or loci that are identified as "weird" in the subsequent analyses. In ipyrad this can be done by searching the .loci file, in dDocent... If the alignment looks wonky, you can feel better about removing the whole locus (?).  
<!-- `r colorize("code example .loci", "red")` -->
<!-- `r colorize("code dDocent", "red")`   -->
<!-- `r colorize("code stacks", "red")`     -->

@@

<!-- `r colorize("PCA figure of batch effect", "red")` -->

You can identify which loci are driving the batch effect by 1) identifying the top PC loadings (SNPs) and/or 2) do an Fst outlier analysis with the batches specified as populations (eg, Bayescan). Both of these methods are most appropriate if you've randomized your samples between batches. If you have not randomized samples but are certain there is a batch effect driving variation in your dataset, you can still remove loci this way but you may also be removing biologically informative loci.

@@

Once you've identified these batch effect SNPs, it can be useful to [visualize the alignments](#align) before removing them. If the entire alignment for that locus looks wonky, you should remove all SNPs from that RAD locus from your final dataset before conducting analyses.     
<!-- `r colorize("code to identify top loading snps", "red")` -->

```{r top loading SNPS}
```

#### Cryptic species/contamination/clones {#cryptic}  
One issue that is particularly common, especially for marine invertebrates, are cryptic, yet highly diverged lineages occurring in the same geographic location. If you are trying to do a landscape/seascape genetics study, cryptic species can dramatically throw off population genetic analyses. A PCA using all the samples can help identify cryptic species if samples are forming clusters that do not align with any other possible factor (eg, geography, sequencing batch).   
<!-- `r colorize("PCA figure of cryptic species", "red")`   -->

@@

Cryptic clones can also dramatically throw off population genetic analyses. In a PCA, clones will often appear as their own very tight cluster, similar to sample replicates. Below the PCA shows x and a simple calculate of pairwise-distance (i.e. here manhattan distance) shows x known duplicates, and x unknown clones. *other ways to tell they are clones*  
<!-- `r colorize("PCA figure of cryptic clones", "red")` -->

#### "Bad" samples {#bad}  
Sometimes a sample doesn't sequence well (few sequencing reads, higher error rate). This can be due to DNA quality, an issue during library prep, or not enough sequencing depth (average # of reads per sample). Generally, it will lead to a sample with fewer sequencing reads, higher missing data in a SNP dataset, and fewer shared loci with other samples. Identifying and then removing these samples *prior* to the final RADseq assembly analysis can help minimize mis-assembled loci, genotyping errors, and excessive filtering of acceptable loci.

>>>>

```{r,message = FALSE,echo = FALSE}
  library("knitcitations")
  library("knitr")
  cite_options(citation_format = "pandoc", max.names = 3, style = "html", hyperlink = "to.doc")
  bib <- read.bibtex("bibtexlib.bib")
  opts_chunk$set(tidy = FALSE, message = FALSE, warning = FALSE,
                 fig.width = 10, fig.height = 6, cache = TRUE)
  # use this to set knitr options:
  # http://yihui.name/knitr/options #chunk_options
```
---
title: 'RADseq'
subtitle: '*NJ Gr&uuml;nwald, ZN Kamvar, and SE Everhart*'
---

@@

colorize <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{%s}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'>%s</span>", color,
      x)
  } else x
}
#set bash code chunks to use bash_profile
knitr::opts_chunk$set(engine.opts = list(bash = "-l"))

If you would like to run the R code examples that are scattered throughout the guide (recommended but not required!), you will need to install some R packages:  
```{r, eval=F}
install.packages("tidyverse")
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("SeqArray")
BiocManager::install("SNPRelate")
```

Now load those packages, if using:  
```{r}
library(ggplot2)
library(SeqArray)
library(tidyverse)
library(SNPRelate)
```

We also provide alternative coding examples that are based only on the command line, requiring these software packages:

* vcftools(link)
* gnuplot

# Reduced Representation Sequencing (RADseq/GBS)

"Restriction-site   Associated   DNA   sequencing"  - RADseq - combines  restriction  enzymes  with next-gen, massively parallel,  short-read  sequencing. RADseq involves the use of restriction enzymes which are used to shear DNA at restriction enzyme cutsite. RADseq comes in different flavors. Double-digest RADseq (ddRAD; Peterson et al.,2012)  selects  markers  with  two  restriction enzymes  with  different  cut  frequencies. Genotype-by-Sequencing  (GBS;  Elshire  et  al.,  2011)  uses  a  frequent-cutting  restriction  enzyme  with PCR size selection. See this review for a comparison of some of the more popular protocols (link). In this guide we use the term "RADseq" to refer to any of these protocols, including those that don't involve random shearing of data. When recommendations are specific to a certain type of RAD/GBS, we will explicitly say so.

Some aspects of RADseq data that are unique compared to whole genome resequencing (more):

@@

![Table 1 from O'Leary et al. 2018.](images/oleary_table1.png)

@@

* If this RAD/GBS method has not been done in your species or in your molecular lab setup before, spend some time optimizing the protocol using a representative subset of individuals. Then try to keep everything about the library prep as consistent as possible across samples (eg, DNA extraction kit, PCR cycles, sequencing platform). This isn't always possible, especially if you need to optimize the protocol for certain tricky samples. Just make sure to keep track of everything!   
* Randomize samples across library prep batches and sequencing lanes! For example, if you are sequencing two different groups of samples on two different sequencing lanes, make sure they are randomized in respect to sample location or whatever your groups of interest are. If you are preparing groups of individuals in different batches to be pooled later, randomize across these batches. Alicia MY even recommends randomizing across DNA extractions, especially if the person doing the extractions is new to molecular work. This is to allow you to control for potential [batch effects](#batch) that are often observed with RAD data.  
```{r, eval = F, include=FALSE}
# This code reads in a comma-delimited STRATA/meta-data file, randomises samples, and returns a file with wells and plates and duplicates which may be used to assist with plating libraries
# ** TODO** will need to consider where to duplicate samples - need to consider STARTA, DOC, ect..?
# maybe have user write in a list of samples to replicate

@@

### First, look at the raw data! {#fastqc}  
Always look at your data with fastqc(link) before starting an assembly. First, this is a good check to just make sure the sequencing worked. If you have demultiplexed data, you can use multiqc(link) to generate fastqc plots for all individuals and quickly identify ones that did not sequence well.  
Check out [this informative tutorial](https://datacarpentry.org/wrangling-genomics/02-quality-control/index.html) on running fastqc and interpreting the output.  
Do you have a lot of adapter sequences? Are the ends of your reads very low quality? If so, you should expect a fair amount of trimming and read filtering to occur prior to assembly. If that doesn't occur or if too many reads are being filtered so as to only recover a small number of SNPs, something might need to be tweaked with your trimming and filtering step.

@@

### Run an assembly pipeline {#assembly}

Some groups have also developed pipelines for specific flavors of lab (e.g., Matz lab and 2bRAD) or proprietary software (eg, DARTseq), which we touch on below. **In most cases, it is recommended that you use a pipeline developed and tested for RAD data, especially if you are making a de novo assembly.** LaCava et al, 2019 have an excellent study where they review various de novo assemblers that are used in these pipelines. Assemblers that were not explicitly developed for short reads (eg, Velvet, ABySS) performed very poorly, while CD-HIT (the assembler in dDocent) performed the best. Stacks/Stacks2 and vsearch (the assembler in ipyrad) both performed worse when analyzing simulated data with indels. It should be noted that all of these RAD pipelines have optimized the specific parameters of these de novo assemblers to work with RAD data of various flavors, which may not be fully reflected in the LaCava study. Based on the MarineOmics RAD panel, the general consensus is that all the popular pipelines can produce adequately accurate datasets (with appropriate parameter optimization and data filtering).     
![Table 3 from LaCava et al 2019. CD-HIT is used in dDocent, VSEARCH is used in ipyrad.](images/lacavaTable3.png)

#### iPyrad walkthrough

@@

Once you have processed your samples (or a subset of samples) through a genotyping pipeline, you will have a bunch of different output file options for genetic data. The VCF file is one of the most popular file formats, and is the most versatile for initial data exploration as many programs exist to filter and accept VCF files. Below are a list of potential confounding factors that may exist in your data, and how to tease them out.

**Using SeqArray in R** 
If you are following along with the R code examples, we need to 1st read in our data (skip this section if not running R code).

For the code examples in this section, we primarily make use of the R package SeqArray [Zheng et al., 2017](https://doi.org/10.1093/bioinformatics/btx145), which can read and manipulate VCFs. If you’re familiar with R, SeqArray is simple to use. The package can efficiently store sequence variant calls with hundreds of thousands of individuals,  variants (e.g., SNVs, indels and structural variation calls) and annotations.  It employs a data container called a CoreArray Genomic Data Structure (GDS). It’s super-fast (5X faster than PLINK v1.9; 16X faster than vcftools) and it integrates well with other R packages you might use in your analysis pipeline. (i.e. SNPRelate, SeqVarTools). We also like it because you can filter your data before running certain analyses without generating a separate filtered VCF file.

First, we need to convert our VCF file into the GDS format. We will do this once here, then use the GDS file for subsequent code examples.
```{r}
filename = "OL_subset" #replace with your file name 
filename.gds = paste0("RAD_data/", paste0(filename, ".gds"))
filename.vcf = paste0("RAD_data/", paste0(filename, ".vcf"))
  # 1 . Convert VCF to GDS
SeqArray::seqVCF2GDS(vcf.fn = filename.vcf, out.fn = filename.gds, storage.option="ZIP_RA")
gdsin = SeqArray::seqOpen(filename.gds)
print(paste0("The number of SAMPLES in data: ", length(c(SeqArray::seqGetData(gdsin, "sample.id")))))
print(paste0("The number of SNPs in data: ",  length(c(SeqArray::seqGetData(gdsin, "variant.id")))))
```
It is always helpful to have a metadata file with information for each sample, such as sampling site, sequencing library, etc. In our example, our matadata file (OL.popmap) is tab-delimited and has the column headers:  
ID: sample ID
STRATA: sampling location/population
PLATE: sequencing batch  
Next we read in our metadata file, and make sure the samples are in the same order as your VCF file:  
```{r}
metafile = "RAD_data/OL.popmap"
sample.ids = seqGetData(gdsin, "sample.id")
sample.strata =  read.table(metafile, header = T, sep = "\t") %>%
                  dplyr::select(ID, STRATA, PLATE)
```
Now, on to evaluating our data!

#### "Bad" samples {#bad}

Sometimes a sample doesn't sequence well (few sequencing reads, higher error rate). This can be due to DNA quality, an issue during library prep, or not enough sequencing depth (average # of reads per sample). Generally, it will lead to a sample with fewer sequencing reads, higher missing data in a SNP dataset, and fewer shared loci with other samples. Identifying and then removing these samples *prior* to the final RADseq assembly analysis can help minimize mis-assembled loci, genotyping errors, and excessive filtering of acceptable loci.

1) For every SNP dataset you generate, it is a good idea is always evaluate the missingness per sample (and report this distribution in your manuscript!). Identify samples with way more missingness than the rest, and observe how they look in a PCA and locus sharing plot. If they stick out or all cluster together in the middle, then try removing them from the assembly and seeing if it changes downstream analyses. If so, you may want to specify a missingness cutoff for including samples in the final analysis.

**Missingness in R with SeqArray**
```{r}
#using previously loaded gdsin object
print("Per variant: ")
summary(m1 <- SeqArray::seqMissing(gdsin, per.variant=TRUE))
print("Per sample: ")
summary(m2 <- SeqArray::seqMissing(gdsin, per.variant=FALSE))
samples <- SeqArray::seqGetData(gdsin, "sample.id")
cbind(samples,m2)[order(-m2),]
#plot histogram 
hist(m2,breaks=50)

# close the GDS file
seqClose(gdsin)
```

Another common method of filtering and evaluating you data without ever using R is VCFtools(link). 
**Missingness with vcftools, on the command line**  
```{bash, eval = T}
vcftools --vcf RAD_data/OL_subset.vcf --missing-indv --out RAD_data/OL_subset
# sort the file by most missing data and print the top 10 samples  
cat RAD_data/OL_subset.imiss | (read h; echo "$h"; sort -k5 -r) 
```
You can open the file analysis/OL_subset.imiss in any text editor an look at the missingness values by eye.

To plot the missingness on the command line, you can use gnuplot:
```{bash}
#code from Jon Puritz
mawk '!/IN/' RAD_data/OL_subset.imiss | cut -f5 > totalmissing
gnuplot << \EOF
set terminal dumb size 120, 30
set autoscale
unset label
set title "Histogram of % missing data per individual"
set ylabel "Number of Occurrences"
set xlabel "% of missing data"
#set yr [0:100000]
binwidth=0.01
bin(x,width)=width*floor(x/width) + binwidth/2.0
plot 'totalmissing' using (bin($1,binwidth)):(1.0) smooth freq with boxes
pause -1
EOF
```

It looks like samples X,Y,Z have a lot of missing data relative to other samples. We can remove them from our dataset for now, and just keep exploring our data, but if you decide to exclude them permanently from the analysis you should eventually rerun your genotyping pipeline without those samples, esp. if doing a de novo assembly.

2) Another great way to explore missingness with your data is to construct a heatmap of loci that are genoptyped between pairs of samples. Generally, samples that are more closely related will share more loci with each other due to cutsite dropout (link). Deviations from that pattern can pinpoint bad samples as well as batch effects.
  + [ipyrad API tutorial](https://ipyrad.readthedocs.io/en/latest/API-analysis/cookbook-sharing.html) for making a locus sharing heatmap, requires ipyrad output or ipyrad's vcf2hdf5 conversion  
  + [code to generate locus sharing/missingness heatmap from a VCF](https://github.com/atcg/clustOpt)

@@

One of the most powerful methods for exploring your data is a Principle Components Analysis. From [this genetics tutorial](https://comppopgenworkshop2019.readthedocs.io/en/latest/contents/03_pca/pca.html): "To understand how PCA works, consider a single individual and its representation by its 593,124 markers. Formally, each individual is a point in a 593,124-dimensional space, where each dimension can take only the three possible genotypes indicated above, or have missing data. To visualize this high-dimensional dataset, we would like to project it down to two dimensions. But as there are many ways to project the shadow of a three-dimensional object on a two dimensional plane, there are many (and even more) ways to project a 593,124-dimensional cloud of points to two dimensions. What PCA does is figuring out the “best” way to do this project in order to visualise the major components of variance in the data." See [here](https://michaelaalcorn.medium.com/yet-another-pca-explanation-727dff6ce26a) for a linear algebra-based explanation of PCA).

When you have a VCF file with SNPs, use PCA before extensive filtering or playing with parameters to look at the data. Check which SNPs are associated with axes showing the most variation.

Here we use SeqArray and SNPRelate to run a PCA in R. Other popular methods for PCA include pcangsd(link), ...., but note that ... requires some file conversions everytime you filter.

*Reminder:* Missing data is a feature of RAD. Be aware of how different analysis tools deal with missingness, especially PCA which will fill in all missing data with some values. Here, the PCA from SNPRelate imputes missing genotypes as the dosage mean.

##### Doing a PCA with SeqArray and SNPRelate in R.

```{r}
#open the gds file previously created
filename.gds = "RAD_data/OL_subset.gds" #replace with your file name 
gdsin = SeqArray::seqOpen(filename.gds)
# exclude samples previously identified as having too much missin g data
bad_samples = c("OR3_13_C3","CA4_12_C1","OR3_3_C7","CA4_13_C1","BC2_6_C2","BC2_9_C4")
sample.ids = seqGetData(gdsin, "sample.id")
keep = sample.ids[which(!sample.ids %in% bad_samples)]

```
Whether your data are denovo or reference-based, it is important to filter out linked sites before performing a PCA. Here's how to do it in SeqArray. Note, if you have already filtered your VCF file to have only 1 SNP per RAD locus, you don't need to do this: 
```{r,results='hide'}
snpset <- SNPRelate::snpgdsLDpruning(gdsin, ld.threshold=0.2, autosome.only = F, start.pos="random", num.thread=1, remove.monosnp = T, sample.id = keep)  
snpset.id <- unlist(unname(snpset))
```

Now we will actually run the PCA, again removing the samples with missing data, keeping unlinked SNPs, and removing SNPs with less than 5% minor allele frequency. 
```{r}
# PCA only on SNPs with a minor allele freq greater than 2.5%
pca.out = SNPRelate::snpgdsPCA(autosome.only = F, gdsin, num.thread=2, remove.monosnp = T, maf = 0.05,
                               snp.id=snpset.id,
                               sample.id = keep) # filtering for pruned SNPs
#close the gds file (saves memory)
#seqClose(gdsin)

eig = pca.out$eigenval[!is.na(pca.out$eigenval)]
barplot(100*eig/sum(eig), main="PCA Eigenvalues")
```

Data Info: Below we use RAD data from an Australian marine species complex.
```{r, eval =F}

## The following code will read from a text file containing a list of the VCF files you wish to explore (one file per line) 
## Each line must contain the full file path 
## The code will convert your VCF to GDS
## Perform PCA
## Plot the results

list_vcfs = read.table("analysis/Pop_Map/vcf.list")[[1]]
lapply(list(list_vcfs), function(x) { 
  filename = gsub(".vcf", "", gsub("data/", "", x)) ## NOTE:: you want a pattern that will remove the file path 
  print(paste0("PCA for ... ", filename))
  filename.gds = paste0("data/gds/", paste0(filename, ".gds"))
  gdsfmt::showfile.gds(closeall=TRUE)
  # 1 . Convert VCF to GDS
  SeqArray::seqVCF2GDS(vcf.fn = x, out.fn = filename.gds, storage.option="ZIP_RA")
  gdsin = SeqArray::seqOpen(filename.gds)
  print(paste0("The number of SAMPLES in data: ", length(c(SeqArray::seqGetData(gdsin, "sample.id")))))
  print(paste0("The number of SNPs in data: ",  length(c(SeqArray::seqGetData(gdsin, "variant.id")))))
}) # end apply

sample.ids = seqGetData(gdsin, "sample.id")
sample.strata =  read.table("data/test.metadata.csv", header = T, sep = "\t") %>%
                  dplyr::select(TARGET_ID, STRATA, LONG, LAT, PLATE)
id.order = sapply(sample.ids, function(x,df){which(df$TARGET_ID == x)}, df=sample.strata) #incase your strata file is not in the same order as your vcf
sample.strata.order = sample.strata[id.order,]

@@

First color/shape the individuals by the factor you expect to matter (eg, sampling site or region, family, ecotype).  
```{r, eval=T}
#PLOT PCA
#PC1 v PC2 colored by collection location
id.order = sapply(keep, function(x,df){which(df$ID == x)}, df=sample.strata) #incase your strata file is not in the same order as your vcf
sample.strata.order = sample.strata[id.order,]
print(
  as.data.frame(pca.out$eigenvect) %>%
      tibble::add_column(., STRATA =  sample.strata.order$STRATA) %>%
      ggplot(., aes(x=V1, y=V2, color = STRATA)) + 
      geom_point(size=2) +
      stat_ellipse(level = 0.95, size = 1) +
      geom_hline(yintercept = 0) +
      geom_vline(xintercept = 0) +
      theme_bw() +
      xlab(paste0("PC1 [",paste0(round(eig[1], 2)), "%]")) +
      ylab(paste0("PC2 [",paste0(round(eig[2], 2)), "%]")) +
      ggtitle("PCA Colored by Collection Location")
)
```

#### Questions to ask about your PCA: {.unlisted .unnumbered}

* Is there any clustering? If so, is it different than your expectation?
  + If different than expectation, suggests either [batch effects](#batch), [cryptic variation](#cryptic) among individuals, or systematic biological issues affecting assembly/mapping
* Are there outlier samples driving a large amount of variation?
  + If so, they may be [cryptic species](#cryptic), clones/sample replicates, or they [sequenced/genotyped poorly](#bad)
  + In our example we see 2 potential outlier samples in the bottom left corner. We can check which samples they are by plotting the PCA with sample ID and/or looking at the actual PC scores.

```{r, eval=T}
#PLOT PCA
#PC1 v PC2 colored by collection location
print(
  as.data.frame(pca.out$eigenvect) %>%
      tibble::add_column(., ID =  sample.strata.order$ID) %>%
      ggplot(., aes(x=V1, y=V2, label = ID)) +
      geom_text(size =1.5) +
      geom_hline(yintercept = 0) +
      geom_vline(xintercept = 0) +
      theme_bw() +
      xlab(paste0("PC1 [",paste0(round(eig[1], 2)), "%]")) +
      ylab(paste0("PC2 [",paste0(round(eig[2], 2)), "%]")) +
      ggtitle("PCA with Labels")
)
```

```{r}
# which samples have PC1 > -0.7 and PC2 < -0.3?
as.data.frame(pca.out$eigenvect) %>%
      tibble::add_column(., ID =  sample.strata.order$ID) %>%
  filter(V1 > 0.7 & V2 < -0.3) %>%
  select(ID,V1,V2)
```

* Is there no clustering at all? If so, it may be due to:  
  + bioinformatic artifacts leading to noise, [see Testing Parameters](#test)
  + issues with how you are [filtering the data](#filter) (eg, too strict or too lax)
  + issues with how your PCA is treating missing data
  + or you actually have no structure in your dataset!
  + `r colorize("figure example, no clustering", "red")`

@@

Here is an example of a PCA with sample replicates, and then with sample replicates removed:

When removing a sample replicate, you can choose one at random or pick the replicate with the least missing data(link). 
```{r, eval=FALSE}
# example code for removing samples from a SNPRelate PCA 
# assumes you have already loaded a vcf file as a GDS
samples.to.remove = c("SampleA_rep","SampleB_rep")
sample.ids = seqGetData(gdsin, "sample.id")
keep = sample.ids[which(!sample.ids %in% bad_samples)]

# PCA only on SNPs with a minor allele freq greater than 2.5%, only keeping samples in keep
pca.out.noreps= SNPRelate::snpgdsPCA(autosome.only = F, gdsin, num.thread=1, remove.monosnp = T, maf = 0.05,
                               snp.id=snpset.id,
                               sample.id = keep)
#close gds file (saves memory)

Next, color the individuals by potential sources of batch effects (sequencing lane, library prep batch, age of tissue sample, person doing the DNA extraction...).    
```{r, eval=T}
# again, but colored by batch
print(
  as.data.frame(pca.out$eigenvect) %>%
      tibble::add_column(., PLATE =  as.factor(sample.strata.order$PLATE)) %>%
      ggplot(., aes(x=V1, y=V2, color = PLATE))  + #label = ID
      geom_point(size=2) +
      stat_ellipse(level = 0.95, size = 1) +
      geom_hline(yintercept = 0) +
      geom_vline(xintercept = 0) +
      theme_bw() +
      xlab(paste0("PC1 [",paste0(round(eig[1], 2)), "%]")) +
      ylab(paste0("PC2 [",paste0(round(eig[2], 2)), "%]")) +
      ggtitle("PCA colored by Batch/Sequencing Plate")
)

@@

#### Visually inspect alignments {#align}  
Another useful approach is to visually inspect alignments for SNPs or loci that are identified as "weird" in the subsequent analyses. In ipyrad this can be done by searching the .loci file, in dDocent... If the alignment looks wonky, you can feel better about removing the whole locus (?).  
`r colorize("code example .loci", "red")`
`r colorize("code dDocent", "red")`  
`r colorize("code stacks", "red")`

@@

@@

`r colorize("PCA figure of batch effect", "red")`

You can identify which loci are driving the batch effect by 1) identifying the top SNPs contributing to PC variation (termed "loadings") and/or 2) do an Fst outlier analysis with the batches specified as populations (eg, Bayescan). Both of these methods are most appropriate if you've randomized your samples between batches. If you have not randomized samples but are certain there is a batch effect driving variation in your dataset, you can still remove loci this way but you may also be removing biologically informative loci.

Let's pretend like we have batch effects that were observed on PC 1. With SNPRelate we would extract the PC loadings for each SNP at each PC, then 
```{r}

SnpLoad <- snpgdsPCASNPLoading(pca.out, gdsin)
#load.df <- cbind(SnpLoad$snp.id,SnpLoad$snploading)
#colnames(load.df) <- c("Snp.ID",)

Once you've identified these batch effect SNPs, it can be useful to [visualize the alignments](#align) before removing them. If the entire alignment for that locus looks wonky, you should remove all SNPs from that RAD locus from your final dataset before conducting analyses.

#### Cryptic species/contamination/clones {#cryptic}  
One issue that is particularly common, especially for marine invertebrates, are cryptic, yet highly diverged lineages occurring in the same geographic location. If you are trying to do a landscape/seascape genetics study, cryptic species can dramatically throw off population genetic analyses. A PCA using all the samples can help identify cryptic species if samples are forming clusters that do not align with any other possible factor (eg, geography, sequencing batch).   
`r colorize("PCA figure of cryptic species", "red")`

@@

Cryptic clones can also dramatically throw off population genetic analyses. In a PCA, clones will often appear as their own very tight cluster, similar to sample replicates. Below the PCA shows x and a simple calculate of pairwise-distance (i.e. here manhattan distance) shows x known duplicates, and x unknown clones. *other ways to tell they are clones*  
`r colorize("PCA figure of cryptic clones", "red")`
